---
title: "Reproducible research Peer Assessment 1"
author: "Alfred Aita"
date: "August 12, 2015"
output: html_document
---
```{r,echo=F}
library(data.table)
```

First we will unzip and load the into a dataframe. And examine it

```{r,echo=TRUE}
Atbl<-read.csv(unzip("activity.zip"),colClasses = c("numeric","Date","numeric"))
Atbl$date<-as.Date(Atbl$date,format = "%Y-%m-%d")
#Atbl$interval<-as.factor(Atbl$interval)
str(Atbl)
head(Atbl,3)
tail(Atbl,3)
```

We now calculate the total number of steps taken,removing Na's

```{r,echo=TRUE}

total.steps<-sum(Atbl$steps,na.rm = TRUE)
total.steps
```

Calculate total steps per day.
```{r }
##x2<-aggregate(Atbl$steps,by=list((substr(Atbl$date,1,61))),sum, na.rm = T)
x2<-aggregate(steps~date,Atbl,sum)
names(x2)<-c("Date",'daily.steps' )
head(x2)
str(x2)
```
plot a histogram  of steps by each day


```{r,echo=TRUE}


hist(x2$daily.steps,col = "yellow",breaks=25,main = "Steps taken per day",
   xlab = "Number of Steps per day",ylab = "frequency per day")  
```


Calculate the median and the mean for steps taken each day
```{r,echo=TRUE}
x3<-aggregate(Atbl$steps,by=list((substr(Atbl$date,1,61))),mean)
meansteps <-round(mean(x2$daily.steps,na.rm = T))

mediansteps <-median(x2$daily.steps,na.rm = T)
names(x3)<-c("Dates","mean.steps.")

x4<-aggregate(Atbl$steps,by=list((substr(Atbl$date,1,61))),median)

```
The Mean steps per day is
```{r,echo= F}
meansteps
```
The Median steps per day is 
```{r,echo= F}
mediansteps
```

## Average daily avtivity pattern
```{r ,echo=T}
x3a<-aggregate(Atbl$steps,by = list(interval=Atbl$interval),mean,na.rm=T)
colnames(x3a)<-c("interval","avg_steps")
#plot(x3a,type="l")
#using ggplot2 to avoid side effects from above
library(ggplot2)
ggplot(x3a, aes(x=as.numeric(interval), y=avg_steps)) +  geom_line(color="orange",size =1)+labs(title="Avg daily activity pattern") +
   theme_bw()

```

find the interval with the most steps

```{r ,echo =T}
maxInterval<-x3a[which.max(x3a$avg_steps),]

```
The interval with the most steps is
```{r echo = T}
maxInterval$interval
```

The count of steps for that interval is
```{r echo =T}
maxInterval$avg_steps

```
## Impute missing values

1. Caculate total missing values from original data set
 ```{r, echo =T}
 missing_vals<-sum(is.na(Atbl$steps))
 
 ```
 
 Total number of missing values is `r missing_vals`. 
 
##Filling missing values
we will use the mean value accross the days at the same interval

```{r,echo=t}
Atbl_f<-Atbl
L <- dim(Atbl_f[1])
 for(i in 1:L) {
   if(is.na(Atbl_f$steps[i])){
     ival<-Atbl_f$interval[i]
     imp<-x3a[x3a$interval==ival,"avg_steps"]
     Atbl_f$steps[i]<-imp}}
  str(Atbl_f)
  
```

Missing values in new data set
```{r, echo=T}
sum(is.na(Atbl_f$steps))
```
Zero output tells us there are no missing values

# Histogram of steps taken per day all values filled.
```{r ,echo =T}
Filled_Steps_Day  <-aggregate(steps~date,Atbl_f,sum)
names(Filled_Steps_Day)<-c("Date",'daily.steps' )

str(Filled_Steps_Day)
head(x2,2)
head(Filled_Steps_Day,2)
```
```{r,echo=TRUE}
hist(Filled_Steps_Day$daily.steps,col = "yellow",breaks= 25,main = "Steps taken per day filled",
   xlab = "Number of Steps per day",ylab = "frequency per day")
```   
Mean steps per day filled
```{r,echo= T}
NewMean<-mean(Filled_Steps_Day$daily.steps,na.rm = T)
NewMean
```
Median steps filled per day
```{r, echo= T }
NewMedian <-median(Filled_Steps_Day$daily.steps,na.rm= T)
NewMedian
```
The Difference between the values of the mean and median are small

  *Before filling e missing values
  
   1. Mean: `r meansteps`
   
   2. Median: `r mediansteps`
   
  *After filling missing values
  
   1. Mean: 
```{r,echo =T}
NewMean
```
   
   2. Median: 
```{r,echo = T} 
NewMedian
```  

#are there differences in activity patterns weekends vs. weekdays

we are using the data with filled in missing values


```{r,echo=T}
weekdays_steps<-function(data){
 weekdays_steps<- aggregate(data$steps,by=list(interval=data$interval),
                FUN= mean,na.rm=T)
   

          colnames(weekdays_steps) <- c("interval","steps")
            weekdays_steps
}
 data_by_weekdays <- function(data) {
   data$weekday <-as.factor(weekdays(data$date))
   weekend_data<-subset(data,weekday %in% c("Saturday","Sunday"))
   weekday_data<-subset(data,!weekday %in% c("Saturday","Sunday"))
    
   weekend_steps <- weekdays_steps(weekend_data)
   weekday_steps <- weekdays_steps(weekday_data)
   
   weekend_steps$dayofweek <- rep("weekend",nrow(weekend_steps))
   weekday_steps$dayofweek <- rep("weekday",nrow(weekday_steps))
   
   data_by_weekdays<- rbind(weekend_steps,weekday_steps)
   data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
   data_by_weekdays
 }
 data_weekdays <- data_by_weekdays(Atbl_f)
```

```{r,echo=T}
 ggplot(data_weekdays,aes(x=interval,y=steps))+
          geom_line(color="red") +
          facet_wrap(~dayofweek,nrow = 2,ncol = 1)+
          labs(x="interval",y="number of steps")+
          theme_bw()
```
        